Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-25 manuscript.rmd
1. Changed isotope data label 2. Removed italics for Psi and delta

\
\     Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies. Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically plant's ability to acquire water via differing rooting depths in terms of hydrological rather than meteorological droughts, are rarely studied, especially in the tropical region. We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.

@@

\
\     In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture. The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot is situated in the Conrad catchment. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

@@

To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site, when available, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles) [Need to fix SI Figs. numbering to begin at 1]). For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\Psi$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

\begin{equation}
\Psi_\textrm{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\Psi_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively. $\Psi_e$, $\theta_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.

@@

We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up. Extreme soil droughts were identified by depth $z$ as days that fell below the 5th percentile of $\Psi_{\textrm{soil},z}$ for a given day of the year.

@@

To estimate effective rooting depth, we tested alternate models for growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many intrinsic and extrinsic factors that limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Rhizosphere soil water potential ($\Psi$~soil~) down-regulates plant tissue water potentials [@Christoffersen:2016kd2d]. In moist forests $\Psi$~leaf~ equilibrates with $\Psi$~soil~ at night and declines through the day as plants transpire. $\Psi$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which in turn controls stomatal conductance and photosynthetic rates [@Sack.2012]. We therefore used maximum diurnal leaf conductance ($K_\textrm{leaf}$) as determined by $\Psi$~soil~ as a proxy for the soil water limitation on growth. In our ERD models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential. This approach does not specifically incorporate the buffering effect of stem water storage on $K_\textrm{leaf}$, which may cause the model to slightly overestimate ERD (see Discussion). VPD has a positive impact on transpiration and thus photosynthesis and growth upto a VPD threshold, after which VPD has a negative impact [@Grossiord.2020]. The fraction of a tree's leaf cover present at a given time further limits it's growth potential. We therefore regressed species-specific growth in a census interval against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

\
\    For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\Psi_{\textrm{soil},s,i|z,h}$in the leaf vulnerability curves defined below in Eq. \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:

\begin{equation}
K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}\Psi_{\textrm{soil},i|z,h}}
(\#eq:k-soil)
\end{equation}
\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\Psi_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yrly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization *h* we retained only one depth that had the maximum likelihood (or, the minimum AIC score), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $R$^2 >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :

\begin{equation}
\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big\{z,h\,\Big|\, \Big(\textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|~z,h})\Big) \cap \Big(r_{s|z,h} \ge 0.71\Big)\Big\}
(\#eq:erd)
\end{equation}
where $L$ is the likelihood, $H$ = {1, .., 100}, refer to the 100 hydrological realisations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m. We removed finely resolved soil layers below 0.21 m, averaged $\Psi_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m.

@@

We obtained leaf hydraulic vulnerability curves ($K_\textrm{leaf}$ vs. $\Psi$~leaf~; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al. (to be submitted; see SI for brief description of methods) described as:

@@

\
\   As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq. \#ref(eq:vc-exp). For all the `r length(erd.sp)` ERD species we estimated species $\Psi$~50,leaf~ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).

@@

Xylem sap $\mathrm{\delta}$^2^H values can provide an index of rooting depth, which we used to validate the ERD models as an independent observation of ERD. We used xylem sap $\mathrm{\delta}$^2^H values from BCI given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding vertical gradient of $\mathrm{\delta}$^2^H in the soil. (Fig. S\@ref(fig:iso-data)). Briefly, this isotopic approach makes use of a naturally formed gradient in deuterium to hydrogen ratio of soil water vertically belowground, which is generally formed due to evaporative enrichment (here, during the dry season). Shallow soil water has a greater deuterium to hydrogen content than deep soil water (here, $\mathrm{\delta}$^2^H ranged from -16.5 to -50 ‰ over 0.05 to 1 m depth), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water. As root water-uptake is generally a non-fractionating process, $\mathrm{\delta}$^2^H of water in the xylem sap of a tree reflects the signature of the source water. Thus water-sourcing depth of the tree can be identified as the depth with the same value of $\mathrm{\delta}$^2^H as that of the xylem sap, with the uncertainty that the value of $\mathrm{\delta}$^2^H may also result from a mixing of water from shallower or deeper depths [@Dawson.1991]. Analyses of xylem sap $\mathrm{\delta}$^2^H of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water. Two other had water access at least 0.8 m deep, one was restricted to water access between 0 to 0.2 m and five fell in the range over which the relationship between soil depth and soil water $\mathrm{\delta}$^2^H was variable in the 0.2 to 0.7 m depth range.

\   We compared six species with data for $\mathrm{\delta}$^2^H as well as modelled estimates of ERD, after removing those species in @Meinzer:2001 that are likely to have no leaves in March 1997 because xylem sap $\mathrm{\delta}$^2^H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves. We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile. For each model of ERD (Eq. \@ref(eq:M-chosen) and Eqs. in SI), estimated species ERD was regressed against species xylem sap $\mathrm{\delta}$^2^H (‰).

@@

To evaluate whether ERD was associated with above-ground hydraulic traits ($n$ = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_\textrm{max,stem}$), leaf turgor loss point ($\Psi$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi$~88,stem~), and hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~). See [SI](#SIData.brett) for data collection and estimation of these variables.

@@

\
\   We regressed species ERD against species ($\textrm{Mort}_{s,t}$ for each of the census interval $t$). To assess whether deeper ERD species escaped the experience of significant soil drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\Psi_{\textrm{soil},z=\textrm{ERD}}$) were below species $\Psi$~crit~. $\Psi$~crit~ was defined as $\Psi$~50,leaf~.

@@

\   Soil Water Potential dynamics ($\Psi_{\textrm{soil},z}$) by depth $z$ (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig. \@ref(fig:swp-droughts)). Notably, soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years each, depending on depth location (Fig. S\@ref(fig:swp-droughts)). Across the simulation period (1990-2015) $\Psi_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m in the simulations (Fig. S\@ref(fig:swp-all-depths)).

@@

The chosen model (with a mulicative effect of VPD and not LAI; Eq. \@ref(eq:M-chosen)) explained a large fraction of the variance in the xylem sap isotopic signature ($\mathrm{\delta}$^2^H); $R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig. \@ref(fig:erd-iso)). Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models; <20 speceis vs. `r length(erd.sp)` species for the chosen model (Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the less parimonious, chosen model (with VPD and not LAI, Eq. \@ref(eq:M-chosen)).

@@

\
\   In this forest, modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig. \@ref(fig:sp-erd)). Notably, two species, *L. seemannii* and *T. tuberculata*, that @Meinzer:1999 found to have xylem sap $\mathrm{\delta}$^2^H values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig. \@ref(fig:erd-iso)).

@@

For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_\textrm{max,stem}$, Spearmans' $r$ = `r cor.k`, $p$ = `r p.k`; Fig. \@ref(fig:erd-hyd)A), less negative leaf $\Psi$~tlp~ (Spearmans' $r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig. \@ref(fig:erd-hyd)B), less negative $\Psi$~88,stem~ and thus greater vulnerability to embolism from cavitation (Spearmans' $r$ = `r cor.stem.88`, $p$ = `r p.stem.88`; Fig. \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~; Spearmans' $r$ = `r cor.hsm`, $p$ = `r p.hsm`; Fig. \@ref(fig:erd-hyd)D).

@@

Our novel approach for estimating ERD using 25 years of growth, species-specific leaf vulnerability curves, modeled soil moisture through the profile, and observed VPD was consistent with estimates using xylem water isotopes. Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety (Fig. \@ref(fig:erd-hyd)). Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1981-2015) (Fig. S\@ref(fig:erd-mort-census), especially amongst the evergreen species (Fig. \@ref(fig:erd-mort-census-evg)). Analysis of time spent below ($\Psi$~crit~) in the soil zone matching ERD $\Psi_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig. \@ref(fig:erd-time); Fig. S\@ref(fig:erd-time-evg)). We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring large tree species.

@@

\
Figure \@ref(fig:ab-fit): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels). Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\Psi_{\textrm{soil},z}$) for a given DOY and depth $z$ are shown (same across census intervals). $\Psi_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if for some part of that year $\Psi_{\textrm{soil},z}$ declined below the 5 percentile distribution.

>>>>

\
\     Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies. Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically plant's ability to acquire water via differing rooting depths in hydrological rather than meteorological droughts, are rarely studied, especially in the tropical region. We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.

@@

\
\     In the 50 ha plot, all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture. The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot is situated in the Conrad catchment. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

@@

To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site, when available, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles) [Need to fix SI Figs. numbering to begin at 1]). For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

\begin{equation}
\Psi_\textrm{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively. $\mathrm{\Psi}_e$, $\theta_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.

@@

We ran ELM-FATES with the chosen 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up. Extreme soil droughts were identified by depth $z$ as days that fell below the 5th percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.

@@

To estimate effective rooting depth, we tested alternate models for growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many intrinsic and extrinsic factors that limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Rhizosphere soil water potential ($\mathrm{\Psi}$~soil~) down-regulates plant tissue water potentials [@Christoffersen:2016kd2d]. In moist forests $\mathrm{\Psi}$~leaf~ equilibrates with $\mathrm{\Psi}$~soil~ at night and declines through the day as plants transpire. $\mathrm{\Psi}$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which in turn controls stomatal conductance and photosynthetic rates [@Sack.2012]. We therefore used maximum diurnal leaf conductance ($K_\textrm{leaf}$) as determined by $\mathrm{\Psi}$~soil~ as a proxy for the soil water limitation on growth. In our ERD models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential. This approach does not specifically incorporate the buffering effect of stem water storage on $K_\textrm{leaf}$, which may cause the model to slightly overestimate ERD (see Discussion). VPD has a positive impact on transpiration and thus photosynthesis and growth upto a VPD threshold, after which VPD has a negative impact [@Grossiord.2020]. The fraction of a tree's leaf cover present at a given time further limits it's growth potential. We therefore regressed species-specific growth in a census interval against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

\
\    For each species, we estimated diurnal maximum value of $K_{\textrm{leaf},s,i|z,h}$ for a species $s$ by substituting $\mathrm{\Psi}_{\textrm{soil},s,i|z,h}$in the leaf vulnerability curves defined below in Eq. \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:

\begin{equation}
K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}\Psi_{\textrm{soil},i|z,h}}
(\#eq:k-soil)
\end{equation}
\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yrly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization *h* we retained only one depth that had the maximum likelihood (or, the minimum AIC score), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $R$^2 >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :

\begin{equation}
\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big\{z,h\,\Big|\, \Big(\textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|~z,h})\Big) \cap \Big(r_{s|z,h} \ge 0.71\Big)\Big\}
(\#eq:erd)
\end{equation}
where $L$ is the likelihood, $H$ = {1, .., 100}, refer to the 100 hydrological realisations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m. We removed finely resolved soil layers below 0.21 m, averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m.

@@

We obtained leaf hydraulic vulnerability curves ($K_\textrm{leaf}$ vs. $\mathrm{\Psi}$~leaf~; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al. (to be submitted; see SI for brief description of methods) described as:

@@

\
\   As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq. \#ref(eq:vc-exp). For all the `r length(erd.sp)` ERD species we estimated species $\mathrm{\Psi}$~50,leaf~ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).

@@

Xylem sap $\mathrm{\delta}$^2^H values (hereafter, $\mathrm{\delta}$^2^H~xylem~) can provide an index of rooting depth, which we used to validate the ERD models as an independent observation of ERD. We used species $\mathrm{\delta}$^2^H~xylem~ values from BCI given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding vertical gradient of $\mathrm{\delta}$^2^H in the soil (hereafter, $\mathrm{\delta}$^2^H~soil~ ). (Fig. S\@ref(fig:iso-data)). Briefly, this isotopic approach makes use of a naturally formed gradient in deuterium to hydrogen ratio of soil water vertically belowground, which is generally formed due to evaporative enrichment (here, during the dry season). Shallow soil water has a greater deuterium to hydrogen content than deep soil water (here, $\mathrm{\delta}$^2^H~soil~ ranged from -16.5 to -50 ‰ over the depth of 0.5-1 m), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water. As root water-uptake is generally a non-fractionating process, $\mathrm{\delta}$^2^H~xylem~ of a tree reflects the signature of the source water. Thus water-sourcing depth of the tree can be identified as the depth with the same value of $\mathrm{\delta}$^2^H~soil~, or as that of $\mathrm{\delta}$^2^H~xylem~, with the uncertainty that the value of $\mathrm{\delta}$^2^H~xylem~ may also result from a mixing of water from shallower or deeper depths [@Dawson.1991]. Analyses of $\mathrm{\delta}$^2^H~xylem~ of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water. Two other had water access at least 0.8 m deep, one was restricted to water access between 0 to 0.2 m and five fell in the range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable in the 0.2 to 0.7 m depth range.

\   We compared six species with data for $\mathrm{\delta}$^2^H~xylem~ as well as modelled estimates of ERD, after removing those species in @Meinzer:2001 that are likely to have no leaves in March 1997 because xylem sap $\mathrm{\delta}$^2^H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves. We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile. For each model of ERD (Eq. \@ref(eq:M-chosen) and Eqs. in SI), estimated species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~ (‰).

@@

To evaluate whether ERD was associated with above-ground hydraulic traits ($n$ = 8 species with both datasets), we regressed species ERD against maximum stem hydraulic conductance ($K_\textrm{max,stem}$), leaf turgor loss point ($\mathrm{\Psi}$~tlp~), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\mathrm{\Psi}$~88,stem~), and hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~). See [SI](#SIData.brett) for data collection and estimation of these variables.

@@

\
\   We regressed species ERD against species ($\textrm{Mort}_{s,t}$ for each of the census interval $t$). To assess whether deeper ERD species escaped the experience of significant soil drought, we calculated for each species the proportion of days in the simulation period (1990-2015) during which soil water potential in the soil zone matching species ERD ($\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$) were below species $\mathrm{\Psi}$~crit~. $\mathrm{\Psi}$~crit~ was defined as $\mathrm{\Psi}$~50,leaf~.

@@

\   Soil Water Potential dynamics ($\mathrm{\Psi}_{\textrm{soil},z}$) by depth $z$ (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig. \@ref(fig:swp-droughts)). Notably, soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years each, depending on depth location (Fig. S\@ref(fig:swp-droughts)). Across the simulation period (1990-2015) $\mathrm{\Psi}_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m in the simulations (Fig. S\@ref(fig:swp-all-depths)).

@@

The chosen model (with a mulicative effect of VPD and not LAI; Eq. \@ref(eq:M-chosen)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig. \@ref(fig:erd-iso)). Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models; <20 speceis vs. `r length(erd.sp)` species for the chosen model (Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the less parimonious, chosen model (with VPD and not LAI, Eq. \@ref(eq:M-chosen)).

@@

\
\   In this forest, modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig. \@ref(fig:sp-erd)). Notably, two species, *L. seemannii* and *T. tuberculata*, that @Meinzer:1999 found to have $\mathrm{\delta}$^2^H~xylem~ values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig. \@ref(fig:erd-iso)).

@@

For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_\textrm{max,stem}$, Spearmans' $r$ = `r cor.k`, $p$ = `r p.k`; Fig. \@ref(fig:erd-hyd)A), less negative leaf $\mathrm{\Psi}$~tlp~ (Spearmans' $r$ = `r cor.tlp`, $p$ = `r p.tlp`; Fig. \@ref(fig:erd-hyd)B), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to embolism from cavitation (Spearmans' $r$ = `r cor.stem.88`, $p$ = `r p.stem.88`; Fig. \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~; Spearmans' $r$ = `r cor.hsm`, $p$ = `r p.hsm`; Fig. \@ref(fig:erd-hyd)D).

@@

Our novel approach for estimating ERD using 25 years of growth, species-specific leaf vulnerability curves, modeled soil moisture through the profile, and observed VPD was consistent with estimates using xylem water isotopes. Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety (Fig. \@ref(fig:erd-hyd)). Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1981-2015) (Fig. S\@ref(fig:erd-mort-census), especially amongst the evergreen species (Fig. \@ref(fig:erd-mort-census-evg)). Analysis of time spent below ($\mathrm{\Psi}$~crit~) in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig. \@ref(fig:erd-time); Fig. S\@ref(fig:erd-time-evg)). We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring large tree species.

@@

\
Figure \@ref(fig:ab-fit): Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels). Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\mathrm{\Psi}_{\textrm{soil},z}$) for a given DOY and depth $z$ are shown (same across census intervals). $\mathrm{\Psi}_{\textrm{soil},z}$ by DOY for a year (colored line) is only shown if for some part of that year $\mathrm{\Psi}_{\textrm{soil},z}$ declined below the 5 percentile distribution.
